Distance Sampling in \proglang{R}

### 2015-12-13 paper.Rmd
start Tiago changes, need to merge Eric first...

---
author:
  - name: David L Miller
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Eric Rexstad
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: Eric.Rexstad@st-andrews.ac.uk
  - name: Len Thomas
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: len.thomas@st-andrews.ac.uk
  - name: Laura Marshall
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: lhm@st-andrews.ac.uk
  - name: Jeffrey L Laake
    affiliation: NOAA
    address: >
      National Marine Mammal Laboratory
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The estimation of the abundance of biological population is an essential part of applied ecology and wildlife management. To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this. Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}. We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, its interface to other packages to provide sophisticated analyses and a migration pathway from previous software.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, "\\proglang{R}"]
  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations. Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler. We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.

The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT"). From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [@Thomas:2010cf, further details of the history of Distance for Windows are included in that article]. This article shows how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg]. \pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} [@mrds-pkg] and offers a subset of the analyses possible with that package.

>>>>

---
author:
  - name: David L Miller
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Eric Rexstad
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: Eric.Rexstad@st-andrews.ac.uk
  - name: Len Thomas
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: len.thomas@st-andrews.ac.uk
  - name: Laura Marshall
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: lhm@st-andrews.ac.uk
  - name: Jeffrey L Laake
    affiliation: NOAA
    address: >
      National Marine Mammal Laboratory
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The estimation of the abundance of a biological population is an essential part of applied ecology and wildlife management. To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this. Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}. We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, its interface to other packages to provide sophisticated analyses and a migration pathway from previous software.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, "\\proglang{R}", Distance]
  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, R, Distance]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations. Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler. We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.

The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT"). From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [further details of the history of Distance for Windows are included in @Thomas:2010cf]. Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg]. \pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.
